\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-11-04 k-planar.tex
Some little cleanups in the proof of Theorem 7

  Begin with any width-$t$ tree-decomposition $(B_x:x\in V(T_0))$ of $H$ that uses some tree $T_0$ and has no empty bags.  Select any node $x_1\in V(T_0)$, add a leaf $x_0$, with $B_{x_0}=\emptyset$, adjacent to $x_1$ and root $T_0$ at $x_0$. (The purpose of $x_0$ is to ensure that every node $x$ for which $B_x$ is non-empty has a parent.)  Let $f:V(H)\to V(T)$ be the function that maps each $x\in V(H)$ onto the root of the subtree $T_0[x]:=T_0[\{y\in V(T_0): x\in B_y\}]$.  If $f$ is not one-to-one, then select some distinct pair $x,y\in V(H)$ with $a:=f(x)=f(y)$.  Subdivide the edge between $a$ and its parent in $T_0$ by introducing a new node $a'$ with $B_{a'}=B_{a}\setminus\{x\}$. Now $f(y)=a'$ and $f(x)=a$, so this modification reduces the number of distinct pairs $x,y\in V(H)$ with $f(x)=f(y)$.  Repeatedly performing this modification will eventually produce a tree-decomposition $(B_x:x\in V(T_0))$ of $H$ in which $f$ is one-to-one.

@@

For a node $x\in V(T)$, we say that a shortcut $P\in\SS$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$.  In other words, if $P=v_0,\ldots,v_r$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x$ if $v\in Y_x$ or if $\SS$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$.  For each $v\in V(G)$, let $a(v)$ denote the lowest common $T$-ancestor of all the nodes $x\in V(T)$ in which $v$ participates.

@@

  Consider the subgraph $H_v:=H[x\in V(T):\text{$v$ participates in $x$}\}]$.  Since $v\in Y_{x_0}$, $v$ participates in $x_0$, so $x_0\in V(H_v)$.  We claim that, for each $x\in V(H_v)\setminus\{x_0\}$, $H_v$ contains a path $y_0,\ldots,y_s$ of length $s\le k-1$ from $y_0:=x_0$ to $y_s:=x$.  Indeed, since $v$ participates in $x\neq x_0$, $\mathcal{S}$ contains a shortcut $P$ with $v\in V(P)$ and that crosses $x$.  Therefore $P$ contains a path $v_0,\ldots,v_s$ from $v_0:=v$ to some $v_s\in Y_x$.   Furthermore $v_s$ is an internal vertex of $P$ and $P$ has length at most $k$, so $s\le k-1$.  For each $i\in\{0,\ldots,s\}$, let $Y_{y_i}$ be the part of $\mathcal{Y}$ that contains $v_i$.  For each $i\in\{1,\ldots,s\}$, the existence of $v_{i-1}v_{i}\in E(G)$ implies that $y_{i-1}=y_i$ or $y_{i-1}y_i\in E(H)$.
  Therefore $y_0,\ldots,y_s$ is a lazy walk\footnote{A \defin{lazy walk} in a graph $G$ is a walk in the pseudograph obtained by adding a self-loop at each vertex of $G$.} in $H_v$ of length at most $s\le k-1$.  This lazy walk contains a path from $y_0$ to $y_s$ of length at most $s\le k-1$, as promised.

  Since $H_v$ contains a path from $x_0$ to $x$ for each $x\in V(H_v)\setminus\{x_0\}$,  $H_v$ is connected.  Since $H_v$ is connected, $x_r$---which is the lowest common $T$-ancestor of $V(H_v)$---is itself a node of $V(H_v)$.  Therefore $H_v$ contains a path $x_0,\ldots,x_r$ of length most $k-1$.  By definition, $v$ participates in each of $x_0,\ldots,x_r$.
\end{proof}

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Since $a:V(G)\to V(T)$ is a function with domain $V(G)$ and range $V(T)$, $\PP:=\{S_x : x\in V(T)\}$ is a partition of $V(G)$. Let $J:=G^\SS/\PP$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

@@

  \begin{enumerate}
    \item $x_r$ is a strict $T$-ancestor of $x$.  The shortcut $P=x_0,\ldots,x_r$ implies that $x_0$ participates in $x_{r-1}$.  Therefore $x'=a(x_0)$ is a $T$-ancestor of $x_{r-1}$, so $x$ is also a $T$-ancestor of $x_{r-1}$.  Since $x_{r-1}$ is a $T$-descendant of $x$ and $x_{r-1}x_r\in E(H)$, $x_r=y_0$ is contained in $B_x$. Therefore $xy_0$ is an edge of $H^+$, so the path $x,y_0,\ldots,y_s$ is a path in $H^+$ from $x$ to $a$ of length at most $s+1\le k$ and $v_r$ participates in each of $y_0,\ldots,y_s$.

    \item $x_r=x$.  In this case, $y_0,\ldots,y_s$ is a path from $x$ to $a$ of length at most $k-1$ and $v_r$ participates in each of $y_0,\ldots,y_s$.

    \item $x_r$ is a strict $T$-descendant of $x$.
      Since  $y_0$ is a strict $T$-descendant of $x$ and $y_s$ is a $T$-ancestor of $x$, at least one of $y_0,\ldots,y_s$, say $y_i$, is contained in $B_x$. Therefore $xy_i$ is an edge of $H^{+}$.  Therefore $x,y_i,\ldots,y_s$ is a path in $H^+$ from $x$ to $a$ of length at most $s+1\le k$ and $v_r$ participates in each of $y_i,\ldots,y_s$.

  \end{enumerate}
  In each case, we obtain a path $z_0,\ldots,z_{s'}$ of length at most $k$ from $z_0:=x$ to $z_{s'}:=a$ such that $v_r$ participates in each of $z_1,\ldots,z_{s'}$.  By definition, $a$ is a $T$-ancestor of $x$.  Since $v_r$ participates in each of $z_1,\ldots,z_{s'}$, $a=a(v_r)$ is also a $T$-ancestor of each of $z_1,\ldots,z_{s'}$.  Let $W$ be the sequence obtained from $z_0,\ldots,z_{s'}$ by replacing $z_i$ with the lowest common $T$-ancestor of $z_0,\ldots,z_i$, for each $i\in\{0,\ldots,s'\}$.  Then $W$ is a lazy walk in $H^+$ of length at most $k$ that begins at $z_0=x$ and ends at $z_{s'}=a$. By removing duplicates from $W$ we obtain a path in $H^+$ that is also a directed path in $\overrightarrow{H}^+$.  This completes the proof of \cref{general-bag-size}.
\end{proof}

>>>>

  Begin with any width-$t$ tree-decomposition $(B_x:x\in V(T_0))$ of $H$ that uses some tree $T_0$ and has no empty bags.  Select any node $x_1\in V(T_0)$, add a leaf $x_0$, with $B_{x_0}=\emptyset$, adjacent to $x_1$ and root $T_0$ at $x_0$. (The purpose of $x_0$ is to ensure that every node $x$ for which $B_x$ is non-empty has a parent.)  Let $f:V(H)\to V(T)$ be the function that maps each $x\in V(H)$ onto the root of the subtree $T_0[x]:=T_0[\{y\in V(T_0): x\in B_y\}]$.  If $f$ is not one-to-one, then select some distinct pair $x,y\in V(H)$ with $a:=f(x)=f(y)$.  Subdivide the edge between $a$ and its $T$-parent by introducing a new node $a'$ with $B_{a'}=B_{a}\setminus\{x\}$. Now $f(y)=a'$ and $f(x)=a$, so this modification reduces the number of distinct pairs $x,y\in V(H)$ with $f(x)=f(y)$.  Repeatedly performing this modification will eventually produce a tree-decomposition $(B_x:x\in V(T_0))$ of $H$ in which $f$ is one-to-one.

@@

For a node $x\in V(T)$, we say that a shortcut $P\in\SS$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$.  In other words, if $P=v_0,\ldots,v_r$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x\in V(T)$ if $v\in Y_x$ or if $\SS$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$.  For each $v\in V(G)$, let $a(v)$ denote the lowest common $T$-ancestor of all the nodes $x\in V(T)$ in which $v$ participates.

@@

  Consider the subgraph $H_v:=H[\{x\in V(T):\text{$v$ participates in $x$}\}]$.  Since $v\in Y_{x_0}$, $v$ participates in $x_0$, so $x_0\in V(H_v)$.  We claim that, for each $x\in V(H_v)\setminus\{x_0\}$, $H_v$ contains a path $y_0,\ldots,y_s$ of length $s\le k-1$ from $y_0:=x_0$ to $y_s:=x$.  Indeed, since $v$ participates in $x\neq x_0$, $\mathcal{S}$ contains a shortcut $P$ with $v\in V(P)$ and that crosses $x$.  Therefore $P$ contains a path $v_0,\ldots,v_s$ from $v_0:=v$ to some $v_s\in Y_x$.   Furthermore $v_s$ is an internal vertex of $P$ and $P$ has length at most $k$, so $s\le k-1$.  For each $i\in\{0,\ldots,s\}$, let $Y_{y_i}$ be the part of $\mathcal{Y}$ that contains $v_i$.  For each $i\in\{1,\ldots,s\}$, the existence of $v_{i-1}v_{i}\in E(G)$ implies that $y_{i-1}=y_i$ or $y_{i-1}y_i\in E(H)$.
  Therefore $y_0,\ldots,y_s$ is a lazy walk\footnote{A \defin{lazy walk} in a graph $G$ is a walk in the pseudograph obtained by adding a self-loop at each vertex of $G$.} in $H_v$ of length at most $s\le k-1$.  This lazy walk contains a path from $y_0$ to $y_s$ of length at most $s\le k-1$, as promised.

  Since $H_v$ contains a path from $x_0$ to $x$ for each $x\in V(H_v)\setminus\{x_0\}$,  $H_v$ is connected.  Since $H_v$ is connected, $x_r$---which is the lowest common $T$-ancestor of $V(H_v)$---is itself a node of $V(H_v)$.  Therefore $H_v$ contains a path $x_0,\ldots,x_r$ of length most $k-1$.  By the definition of $H_v$, $v$ participates in each of $x_0,\ldots,x_r$.
\end{proof}

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Since $a:V(G)\to V(T)$ is a function with domain $V(G)$ and range $V(T)$, $\PP:=(S_x : x\in V(T))$ is a collection of disjoint sets that cover $V(G)$. Let $J:=G^\SS/\PP$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

@@

  \begin{enumerate}
    \item $x_r$ is a strict $T$-ancestor of $x$.  The shortcut $P=v_0,\ldots,v_r$ implies that $v_0$ participates in $x_{r-1}$.  Therefore $x'=a(v_0)$ is a $T$-ancestor of $x_{r-1}$, so $x$ is also a $T$-ancestor of $x_{r-1}$.  Since $x_{r-1}$ is a $T$-descendant of $x$ and $x_{r-1}x_r\in E(H)$, $x_r=y_0$ is contained in $B_x$. Therefore $xy_0$ is an edge of $H^+$, so the path $x,y_0,\ldots,y_s$ is a path in $H^+$ from $x$ to $a$ of length at most $s+1\le k$ and $v_r$ participates in each of $y_0,\ldots,y_s$.

    \item $x_r=x$.  In this case, $y_0,\ldots,y_s$ is a path from $x$ to $a$ of length at most $k-1\le k$ and $v_r$ participates in each of $y_0,\ldots,y_s$.

    \item $x_r$ is a strict $T$-descendant of $x$.
      Since  $y_0=x_r$ is a strict $T$-descendant of $x$ and $y_s$ is a $T$-ancestor of $x$, at least one of $y_0,\ldots,y_s$, say $y_i$, is contained in $B_x$. Therefore $xy_i$ is an edge of $H^{+}$.  Therefore $x,y_i,\ldots,y_s$ is a path in $H^+$ from $x$ to $a$ of length at most $s+1\le k$ and $v_r$ participates in each of $y_i,\ldots,y_s$.

  \end{enumerate}
  In each case, we obtain a path $z_0,\ldots,z_{s'}$ in $H$ of length at most $k$ from $z_0:=x$ to $z_{s'}:=a$ such that $v_r$ participates in each of $z_1,\ldots,z_{s'}$.  By definition, $a$ is a $T$-ancestor of $x$.  Since $v_r$ participates in each of $z_1,\ldots,z_{s'}$, $a=a(v_r)$ is also a $T$-ancestor of each of $z_1,\ldots,z_{s'}$.  Let $W$ be the sequence obtained from $z_0,\ldots,z_{s'}$ by replacing $z_i$ with the lowest common $T$-ancestor of $z_0,\ldots,z_i$, for each $i\in\{0,\ldots,s'\}$.  Then $W$ is a lazy walk in $H^+$ of length at most $k$ that begins at $z_0=x$ and ends at $z_{s'}=a$. By removing duplicates from $W$ we obtain a path in $H^+$ that is also a directed path in $\overrightarrow{H}^+$.  This completes the proof of \cref{general-bag-size}.
\end{proof}
